Wigglyhedra

### 2024-11-07 FPSAC25/wigglyhedraFPSAC.tex
Wrote up a draft categorical section.

\usepackage{xargs}
\usepackage{hypcap}
\usepackage{paralist}
\usepackage{dsfont}
\usepackage{mathtools}
\usepackage{blkarray}
\usepackage{picins/picins}
\usepackage{overpic}

@@

%% define your authors in the usual way
%% use \addressmark{1}, \addressmark{2} etc for the institutions, and use \thanks{} for contact details
%% for more than 3 names, use the Oxford comma: One Author, Two Author, Red Author, \and Blue Author
\author{Asilata Bapat\thanks{\href{mailto:asilata.bapat@anu.edu.au}{mailto:asilata.bapat@anu.edu.au}. Supported by the Australian DECRA grant DE240100447.}\addressmark{1} \and Vincent Pilaud\thanks{\href{mailto:vincent.pilaud@ub.edu}{mailto:vincent.pilaud@ub.edu}. Supported by the French--Austrian grant PAGCAP (ANR-21-CE48-0020 \& FWF I 5788) and by the Spanish grant PID2022-137283NB-C21.}\addressmark{2}}

@@

\newpage
\section{Categorical representation theory}
\label{sec:representationTheory}

\newpage
%% if you use biblatex then this generates the bibliography
%% if you use some other method then remove this and do it your own way
\printbibliography

>>>>

\usepackage{xargs}
\usepackage{hypcap}
\usepackage{paralist}
\usepackage{dsfont}
\usepackage{mathtools}
\usepackage{blkarray}
\usepackage{picins/picins}
\usepackage{overpic}
\usepackage{tikz-cd}

@@

%% define your authors in the usual way
%% use \addressmark{1}, \addressmark{2} etc for the institutions, and use \thanks{} for contact details
%% for more than 3 names, use the Oxford comma: One Author, Two Author, Red Author, \and Blue Author
\author{Asilata Bapat\thanks{\href{mailto:asilata.bapat@anu.edu.au}{mailto:asilata.bapat@anu.edu.au}. Supported by the Australian grants DE240100447 and DP240101084.}\addressmark{1} \and Vincent Pilaud\thanks{\href{mailto:vincent.pilaud@ub.edu}{mailto:vincent.pilaud@ub.edu}. Supported by the French--Austrian grant PAGCAP (ANR-21-CE48-0020 \& FWF I 5788) and by the Spanish grant PID2022-137283NB-C21.}\addressmark{2}}

@@

\newpage
\section{Categorical representation theory}
\label{sec:representationTheory}
The aim of this section is to interpret the combinatorial results of the previous sections in terms of operations in a triangulated category.
We briefly recall this category, which arises from an algebra associated to the quiver depicted in the following figure.
\begin{figure}[h]
  \centering
  \begin{tikzcd}
    1 \arrow[bend left]{r} & 2 \arrow[bend left]{r}\arrow[bend left]{l}& \cdots \arrow[bend left]{r}\arrow[bend left]{l}& m \arrow[bend left]{l}
  \end{tikzcd}
  \caption{The quiver \(Q_m\). The algebra \(A_m\) is a quotient of the path algebra \(kQ_m\).}
  \label{fig:am-quiver}
\end{figure}

We work over the field \(k = \mathbb{C}\).
Given a quiver \(Q\), we denote by \defn{kQ} its path algebra, which is generated by the idempotent paths \(e_i\) and the arrows \((i,i\pm1)\).
A longer path is denoted \((i_1,\ldots,i_n)\), by the vertices it passes through in order.
\begin{definition}
The \defn{zigzag algebra} \(A_m\) corresponding to the quiver \(Q_m\) depicted in~\cref{fig:am-quiver} is the quotient of \(kQ_m\) by the relations \((i-1, i, i+1) = 0 = (i+1, i, i-1)\) and \((i, i-1, i) = (i, i+1, i)\) for all~\(1 < i < m\).
\end{definition}
The basis elements of \(A_m\) are graded by path-length.
We will consider graded modules over \(A_m\).
Let \(P_i = A_m\cdot e_i\) be the \(i\)th indecomposable projective \(A_m\)-module.
For \(j \in \mathbb{Z}\) and any \(A_m\)-module \(M\), let \(M\langle j \rangle\) denote its graded shift by \(j\).
The triangulated category \(\mathcal{C}_m\) defined below is discussed further in e.g.~\cite[Sect.~2.3.3]{bap.deo.lic:22}.
\begin{definition}
The category \(\mathcal{C}_m\) is the smallest triangulated subcategory of \(D^b(A_m\text{-gr-mod})\) containing all the objects \(P_i\langle j \rangle\).
\end{definition}
Objects in \(\mathcal{C}_m\) are of the form \((M^{\bullet}, d)\), where \(M^{\bullet}\) is a direct sum of graded and homological shifts of objects in \(\{P_1,\ldots, P_m\}\), and \(d \colon M^{\bullet} \to M^{\bullet}\) is a differential that raises homological degree by \(1\).
The \defn{level} of a summand with grading shift \(g\) and homological shift \(h\) is defined to be \(g+h\).
We say that \((M^{\bullet},d)\) is \defn{linear} if each constituent summand has level zero.
We say that an object \((M^{\bullet},d)\) is \defn{spherical} if it has a unique nontrivial endomorphism of graded degree zero, and a unique nontrivial endomorphism of graded degree \(2\), up to scaling.
We refer the reader to~\cite{bel.ber.del:82} for standard results about \(t\)-structures.
\begin{proposition}
  The collection of all linear objects \((M^{\bullet}, d)\) in \(\mathcal{C}_m\) is an abelian subcategory of \(\mathcal{C}_m\), which is the heart of a bounded \(t\)-structure on \(\mathcal{C}_m\).
  Let \(\mathcal{A}_m\) be this abelian subcategory.
\end{proposition}

Let \(\mathcal{C}\) be a triangulated category, and let \(\mathcal{A} \subset \mathcal{C}\) be the heart of a \(t\)-structure on \(\mathcal{C}\).
Then for any integer \(i\), we have the \(i\)th homology functor
\(H^i \colon \mathcal {C} \to \mathcal{A}\),
which truncates an arbitrary object in \(\mathcal{C}\) to its portion in the \(i\)th homological degree.
Furthermore, it is standard that if \(\mathcal{A}\) is the heart of a \emph{bounded} \(t\)-structure, then any object \(X\) has a canonical filtration whose factors are \(H^i(X)[-i]\).
This filtration is called the \defn{cohomology filtration}.

Now consider \(n+2\) points arranged in a horizontal line.
In addition to the wiggly arcs, we also consider \defn{curves}: these are isotopy classes of curves whose endpoints lie on distinct marked points, and which do not cross themselves or any marked point in their interior.
The \emph{Khovanov--Seidel curve model}~\cite[Sect.~4]{kho.sei:02} is a mapping from curves to objects of \(\mathcal{A}_{n+1}\) up to an overall grading and homological shift, as illustrated in~\cref{fig:curveModel}.
An object in \(\mathcal{A}_{n+1}\) lies in the image of this mapping if and only if it is spherical.
Moreover, the minimal intersection number between any pair of curves is the number of morphisms between the associated objects.

\begin{figure}
\centerline{
	\begin{tabular}{c@{\qquad}c@{\qquad}c}
		Spherical objects in bounded 
		&
		&
		\\
		homotopy category of graded projective 
		&
		$\smash{\xrightarrow[\quad\text{combinatorial model}\quad]{}}$
		&
		Curves
		\\
		modules over the zigzag algebra
		&
		&
                \\
		\begin{tikzcd}[ampersand replacement=\&, row sep=-0.1cm, column sep=0.5cm]
			\&           \& P_3 \langle -1 \rangle\arrow{dr} \& \\
			\&           \&                      \& P_2\\
			P_1\langle -5 \rangle\arrow{r}    \& P_1\langle -3 \rangle\arrow{r}   \& P_1 \langle -1 \rangle\arrow{ur} \& 
		\end{tikzcd}
		&
		&
		\raisebox{-.5cm}{\includegraphics[scale=1.3]{curve1}}
		\\[.5cm]
		{\scriptsize cohomology} $\Bigg\downarrow$ {\scriptsize filtration}
		&
		&
		$\Bigg\downarrow$ {\scriptsize cut}
		\\[.5cm]
		Heart of the standard $t$-structure
		&
		$\xrightarrow[\quad\text{combinatorial model}\quad]{}$
		&
		Wiggly arcs
		\\
		\begin{tikzcd}[ampersand replacement=\&, row sep=-0.1cm, column sep=0.5cm]
			P_1\langle -1 \rangle \arrow{dr}\& \\
			\& P_2 \\
			P_3 \langle -1 \rangle \arrow{ur}
		\end{tikzcd}
		&
          	&
		\raisebox{-.5cm}{\includegraphics[scale=1.3]{curve2}}
		\\
		\(P_1\)
		&
		&
		\raisebox{-.5cm}{\includegraphics[scale=1.3]{curve3}	}
	\end{tabular}
}
\caption{An illustration of the Khovanov--Seidel curve model and \cref{prop:cohomology-pieces}.}
\label{fig:curveModel}
\end{figure}
\begin{proposition}
\label{prop:cohomology-pieces}
Let \(\alpha\) be any curve on \(n+2\) points, and let \(X\) be the associated object in \(\mathcal{C}_{n+1}\).
\begin{enumerate}
\item The object \(X[j]\) lies in \(\mathcal{A}_m\) for some \(j \in \mathbb{Z}\) if and only if \(\alpha\) is a wiggly arc.
  In this case, \(X \cong H^j(X)[-j]\), and \(H^i(X) = 0\) for \(i \neq j\).
\item If \(\alpha\) is a general curve and \(j \in \mathbb{Z}\), each indecomposable summand of \(H^j(X)\) is a spherical object of \(\mathcal{A}_m\).
\end{enumerate}
\end{proposition}

\begin{proposition}
\label{prop:hn-filtration}
  Let \(X \in \mathcal{C}_{n+1}\) be a spherical object.
  Consider the set of all indecomposable summands of the nonzero objects \(H^j(X)\).
  Let \(T\) be the corresponding set of wiggly arcs.
  Then
  \begin{enumerate}
  \item 
    \(T\) is a subset of a wiggly pseudotriangulation,
  \item \(T\) contains at most one of the two irrelevant wiggly arcs.
  \end{enumerate}
\end{proposition}
\begin{proposition}
\label{prop:ext1}
Let \(X\) and \(X'\) be objects in the standard heart \(\mathcal{A}_m\) corresponding to two wiggly arcs \(\alpha\) and \(\alpha'\) respectively.
Then the incompatibility degree of \(\alpha\) and \(\alpha'\) is the dimension of \(\Ext_{\mathcal{A}_m}^1(X,X')\).
Thus \(\alpha\) and \(\alpha'\) are compatible (as in~\cref{def:incompatibilityDegree}) if and only if \(\Ext_{\mathcal{A}_m}^1(X, X') = 0\).
\end{proposition}

%\newpage
%% if you use biblatex then this generates the bibliography
%% if you use some other method then remove this and do it your own way
\printbibliography
